Train Random Forest on MAP samples, extract OOB vote probabilities (cluster scores).

```{r}
rm(list=ls())

library(randomForest)
library(dplyr)
library(pROC)
library(ggplot2)

set.seed(14)

# cluster colors: 1=green, 2=blue, 3=red
clus_colors <- c("1" = "#00BA38", "2" = "#619CFF", "3" = "#F8766D")

data_dir <- "/home/zhijiany/workdir/2026/0402_AD_PaperReviews/data/AD"
out_dir <- "/home/zhijiany/workdir/2026/0402_AD_PaperReviews/out/2.RF_score"
```

## Load data

```{r}
countDF <- read.csv(file.path(data_dir, "ROSMAP_AD-N/ROSMAP_AMPAD_AD-N_counts.txt"),
                    sep = '\t', row.names = 1, stringsAsFactors = FALSE)
metaDf <- read.csv(file.path(data_dir, "ROSMAP_AD-N/ROSMAP_AMPAD_AD-N_meta.txt"),
                   sep = '\t', stringsAsFactors = FALSE)
metaDf$newId <- paste0('X', metaDf$newId)

dim(countDF)
table(metaDf$clus)
table(metaDf$study)
```

## Filter to signature genes

```{r}
sig_gene <- readRDS(file.path(data_dir, "gene_order.rds"))
countDF <- countDF[sig_gene, ]
dim(countDF)
```

## Z-scale MAP and ROS separately (using MAP statistics)

```{r}
map_idx <- which(metaDf$study == "MAP")
ros_idx <- which(metaDf$study == "ROS")

map_samples <- metaDf$newId[map_idx]
ros_samples <- metaDf$newId[ros_idx]

# z-scale using MAP statistics
map_counts <- countDF[, map_samples]
ros_counts <- countDF[, ros_samples]

gene_means <- rowMeans(map_counts)
gene_sds   <- apply(map_counts, 1, sd)
gene_sds[gene_sds == 0] <- 1

map_z <- as.data.frame(t((map_counts - gene_means) / gene_sds))
ros_z <- as.data.frame(t((ros_counts - gene_means) / gene_sds))

cat("MAP samples:", nrow(map_z), "\n")
cat("ROS samples:", nrow(ros_z), "\n")
```

---

# Train Random Forest on MAP samples

## Train and OOB evaluation

```{r}
set.seed(14)
train_data <- map_z
train_data$clus <- as.factor(metaDf$clus[map_idx])

mtrySel <- sqrt(ncol(map_z))
rfModel <- randomForest(clus ~ ., data = train_data,
                        importance = TRUE, ntree = 200, mtry = mtrySel)

confusion <- rfModel$confusion
confusion
acc <- (confusion[1, 1] + confusion[2, 2] + confusion[3, 3]) / sum(confusion[, 1:3])
cat("Overall OOB accuracy:", round(acc, 4), "\n")
```

## OOB metrics: per-class accuracy, multiclass AUC

```{r}
# OOB predictions and votes
oob_pred <- rfModel$predicted
oob_votes <- rfModel$votes
true_labels <- as.factor(metaDf$clus[map_idx])

# per-class OOB error rate (from confusion matrix)
cat("Per-class OOB error rates:\n")
for (i in 1:3) {
  cat(sprintf("  Cluster %d: %.4f (accuracy: %.4f)\n",
              i, confusion[i, 4], 1 - confusion[i, 4]))
}

# multiclass AUC (one-vs-rest) using OOB votes
auc_results <- data.frame(class = 1:3, auc = NA)
for (i in 1:3) {
  binary_true <- ifelse(true_labels == i, 1, 0)
  roc_obj <- roc(binary_true, oob_votes[, i], quiet = TRUE)
  auc_results$auc[i] <- auc(roc_obj)
  cat(sprintf("  Cluster %d AUC: %.4f\n", i, auc(roc_obj)))
}
cat(sprintf("  Mean AUC (macro): %.4f\n", mean(auc_results$auc)))
```

## Extract OOB scores (MAP) and predict scores (ROS)

```{r}
# OOB scores for MAP samples
map_scores <- as.data.frame(rfModel$votes)
colnames(map_scores) <- paste0("RF_score_clus", colnames(map_scores))
map_scores$newId <- map_samples
map_scores$RF_predicted_clus <- rfModel$predicted

# Predict on ROS samples
ros_pred_prob  <- predict(rfModel, ros_z, type = "prob")
ros_pred_class <- predict(rfModel, ros_z, type = "class")

ros_scores <- as.data.frame(ros_pred_prob)
colnames(ros_scores) <- paste0("RF_score_clus", colnames(ros_scores))
ros_scores$newId <- ros_samples
ros_scores$RF_predicted_clus <- ros_pred_class

# Combine
all_scores <- rbind(map_scores, ros_scores)
head(all_scores)
```

## Merge scores back into metadata

```{r}
metaDf_scored <- metaDf |>
  left_join(all_scores, by = "newId")

table(Original = metaDf_scored$clus, RF_predicted = metaDf_scored$RF_predicted_clus)

head(metaDf_scored[, c("newId", "study", "clus", "RF_predicted_clus",
                       "RF_score_clus1", "RF_score_clus2", "RF_score_clus3")])
```

---

# OOB confidence distributions (MAP samples)

```{r}
conf_df <- data.frame(
  newId = map_samples,
  study = metaDf$study[map_idx],
  true_clus = as.factor(metaDf$clus[map_idx]),
  predicted_clus = rfModel$predicted,
  max_prob = apply(rfModel$votes, 1, max)
)
conf_df$correct <- conf_df$true_clus == conf_df$predicted_clus

cat("OOB confidence summary (MAP):\n")
conf_df |>
  summarise(
    n = n(),
    mean_confidence = round(mean(max_prob), 4),
    median_confidence = round(median(max_prob), 4),
    accuracy = round(mean(correct), 4)
  )
```

## Confidence distribution by cluster

```{r, fig.width=8, fig.height=5}
p1 <- ggplot(conf_df, aes(x = true_clus, y = max_prob, fill = true_clus)) +
  geom_violin(alpha = 0.6) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.1, size = 1, alpha = 0.4) +
  scale_fill_manual(values = clus_colors) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(title = "RF Prediction Confidence by Cluster (OOB, MAP samples)",
       x = "Cluster", y = "Max Class Probability") +
  theme_bw() +
  theme(legend.position = "none")
print(p1)

ggsave(file.path(out_dir, "confidence_by_cluster_oob.pdf"), p1,
       width = 6, height = 4)
```

## Full probability distribution by true cluster (all 3 scores per sample)

```{r, fig.width=10, fig.height=8}
n_map <- length(map_idx)
votes_long <- data.frame(
  newId = rep(map_samples, 3),
  true_clus = rep(as.factor(metaDf$clus[map_idx]), 3),
  predicted_clus = factor(rep(1:3, each = n_map),
                          labels = paste("Score: Cluster", 1:3)),
  prob = c(rfModel$votes[, 1],
           rfModel$votes[, 2],
           rfModel$votes[, 3])
)

p2 <- ggplot(votes_long, aes(x = predicted_clus, y = prob, fill = predicted_clus)) +
  geom_violin(alpha = 0.6) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.1, size = 0.8, alpha = 0.3) +
  facet_wrap(~ true_clus,
             labeller = labeller(true_clus = function(x) paste("True Cluster", x))) +
  scale_fill_manual(values = setNames(clus_colors, paste("Score: Cluster", 1:3))) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(title = "OOB Vote Probabilities for All Clusters by True Cluster (MAP)",
       x = "", y = "OOB Vote Probability") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1))
print(p2)

ggsave(file.path(out_dir, "votes_by_true_cluster_oob.pdf"), p2,
       width = 10, height = 8)
```

## ROC curves (OOB)

```{r, fig.width=5, fig.height=4.5}
clus_colors <- c("1" = "#00BA38", "2" = "#619CFF", "3" = "#F8766D")

roc_list <- lapply(1:3, function(i) {
  binary_true <- ifelse(true_labels == i, 1, 0)
  roc_obj <- roc(binary_true, oob_votes[, i], quiet = TRUE)
  data.frame(
    sensitivity = roc_obj$sensitivities,
    fpr = 1 - roc_obj$specificities,
    cluster_id = as.character(i),
    auc_label = paste0("Cluster ", i, " (AUC=", round(auc(roc_obj), 3), ")")
  )
})
roc_df <- do.call(rbind, roc_list)

p_roc <- ggplot(roc_df, aes(x = fpr, y = sensitivity, color = cluster_id)) +
  geom_line(linewidth = 0.8) +
  geom_abline(linetype = "dashed", color = "grey50") +
  scale_color_manual(values = clus_colors,
                     labels = unique(roc_df$auc_label)) +
  scale_x_continuous(name = "False Positive Rate") +
  scale_y_continuous(name = "True Positive Rate") +
  labs(title = "ROC Curves — OOB (MAP samples)", color = "") +
  theme_bw() +
  theme(legend.position = "bottom")
print(p_roc)

ggsave(file.path(out_dir, "roc_curves_oob.pdf"), p_roc,
       width = 5, height = 4.5)
```

---

# Summary metrics

```{r}
summary_table <- data.frame(
  Validation = "OOB (MAP-trained model)",
  Accuracy = round(acc, 4),
  Mean_AUC = round(mean(auc_results$auc), 4)
)
print(summary_table)

write.csv(summary_table, file.path(out_dir, "validation_summary.csv"),
          row.names = FALSE)
```

## Save confusion matrix

```{r}
sink(file.path(out_dir, "confusion_matrices.txt"))
cat("=== OOB (MAP-trained model) ===\n")
print(rfModel$confusion)
sink()

write.csv(conf_df, file.path(out_dir, "oob_confidence.csv"),
          row.names = FALSE)

cat("Saved: validation_summary.csv, confusion_matrices.txt, oob_confidence.csv\n")
```

---

# Save results

```{r}
write.table(metaDf_scored,
            file.path(out_dir, "ROSMAP_AMPAD_AD-N_meta_with_RF_scores.txt"),
            sep = '\t', row.names = FALSE, quote = FALSE)

saveRDS(rfModel, file.path(out_dir, "rfModel_clus.rds"))

cat("All outputs saved to:", out_dir, "\n")
```
