
Perform DEG analysis using WM score as a covariate

```{r setup}
rm(list=ls())
library(limma)
library(ggplot2)
library(VennDiagram)

outDir <- "/home/zhijiany/workdir/2026/0402_AD_PaperReviews/out/3.DEG_w_covariate"
dir.create(outDir, recursive = TRUE, showWarnings = FALSE)

dataDir <- "/home/zhijiany/workdir/2026/0402_AD_PaperReviews/data/AD/ROSMAP_AD-N"
rfScoreFile <- "/home/zhijiany/workdir/2026/0402_AD_PaperReviews/out/2.RF_score/ROSMAP_AMPAD_AD-N_meta_with_RF_scores.txt"
```

## Load data

```{r load-data}
# Gene ID to symbol mapping
geneIDNameDf <- read.table(file.path(dataDir, "ensId_symbol_map.txt"),
                           header = TRUE, sep = "\t", stringsAsFactors = FALSE,
                           check.names = FALSE)

# Metadata with RF scores
metaDf <- read.table(rfScoreFile, header = TRUE, sep = "\t",
                     stringsAsFactors = FALSE, check.names = FALSE)
# RF scores file has X-prefixed newId (e.g. X173_120424); strip to match counts columns
metaDf$newId <- sub("^X", "", metaDf$newId)
dim(metaDf)
table(metaDf$diag)
table(metaDf$clus)

# Count matrix
countDf <- read.table(file.path(dataDir, "ROSMAP_AMPAD_AD-N_counts.txt"),
                      header = TRUE, sep = "\t", stringsAsFactors = FALSE,
                      check.names = FALSE)
rownames(countDf) <- countDf[, 1]
countDf <- countDf[, -1]
dim(countDf)
```

## Inspect RF scores

Cluster 2 = WM-related, Clusters 1 & 3 = GM-related.
RF scores sum to 1 per sample, so we use `RF_score_clus2` as the WM score covariate.

```{r inspect-scores}
summary(metaDf[, c("RF_score_clus1", "RF_score_clus2", "RF_score_clus3")])

# Verify scores sum to ~1
range(metaDf$RF_score_clus1 + metaDf$RF_score_clus2 + metaDf$RF_score_clus3)

# Distribution of WM score by diagnosis
ggplot(metaDf, aes(x = diag, y = RF_score_clus2)) +
  geom_boxplot() +
  labs(title = "WM score (RF_score_clus2) by diagnosis", y = "WM score") +
  theme_bw()
```

## Check correlation between diagnosis and WM score

```{r diag-wm-correlation, fig.width=7, fig.height=5}
metaCorrDf <- metaDf
metaCorrDf$diag <- factor(metaCorrDf$diag, levels = c("N", "AD"))
metaCorrDf$diag_num <- ifelse(metaCorrDf$diag == "AD", 1, 0)

pearson_cor <- cor(metaCorrDf$diag_num, metaCorrDf$RF_score_clus2, method = "pearson")
spearman_cor <- cor(metaCorrDf$diag_num, metaCorrDf$RF_score_clus2, method = "spearman")
t_test_res <- t.test(RF_score_clus2 ~ diag, data = metaCorrDf)
wilcox_res <- wilcox.test(RF_score_clus2 ~ diag, data = metaCorrDf)

cat("Diagnosis vs RF_score_clus2 association\n")
cat("  Pearson correlation (diag coded N=0, AD=1):", round(pearson_cor, 4), "\n")
cat("  Spearman correlation (diag coded N=0, AD=1):", round(spearman_cor, 4), "\n")
cat("  Mean RF_score_clus2 in N:", round(mean(metaCorrDf$RF_score_clus2[metaCorrDf$diag == "N"]), 4), "\n")
cat("  Mean RF_score_clus2 in AD:", round(mean(metaCorrDf$RF_score_clus2[metaCorrDf$diag == "AD"]), 4), "\n")
cat("  t-test p-value:", signif(t_test_res$p.value, 4), "\n")
cat("  Wilcoxon p-value:", signif(wilcox_res$p.value, 4), "\n")

p_diag_wm <- ggplot(metaCorrDf, aes(x = diag, y = RF_score_clus2, fill = diag)) +
  geom_boxplot(width = 0.6, alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.12, alpha = 0.5, size = 1.5) +
  scale_fill_manual(values = c("N" = "#7AA6DC", "AD" = "#D97A6C")) +
  labs(
    title = "Association between diagnosis and WM score",
    subtitle = paste0(
      "Pearson r = ", round(pearson_cor, 3),
      "; Spearman rho = ", round(spearman_cor, 3),
      "; t-test p = ", signif(t_test_res$p.value, 3)
    ),
    x = "Diagnosis",
    y = "RF_score_clus2"
  ) +
  theme_bw() +
  theme(legend.position = "none")

ggsave(file.path(outDir, "diag_vs_RF_score_clus2.pdf"), p_diag_wm, width = 6, height = 5)

p_diag_wm
```

## Helper: run limma and save results

```{r helper-functions}
run_limma <- function(countDf, design, coef, geneIDNameDf,
                      padjCut = 0.05, logFCCut = 0.26) {
  fit <- lmFit(countDf, design)
  fit <- eBayes(fit)
  fitRes <- topTable(fit, coef = coef, adjust = "BH", n = Inf)

  resDf <- data.frame(ensid = rownames(fitRes), fitRes, check.names = FALSE)
  resDf <- merge(resDf, geneIDNameDf, by.x = "ensid", by.y = "ensid", all.x = TRUE)
  resDf <- resDf[order(resDf$adj.P.Val), ]

  sig <- resDf[resDf$adj.P.Val <= padjCut, ]
  up <- sig[sig$logFC >= logFCCut, ]
  dn <- sig[sig$logFC <= -logFCCut, ]

  cat("Total genes tested:", nrow(resDf), "\n")
  cat("Significant (adj.P <=", padjCut, "):", nrow(sig), "\n")
  cat("  Up (logFC >=", logFCCut, "):", nrow(up), "\n")
  cat("  Down (logFC <=", -logFCCut, "):", nrow(dn), "\n")

  list(all = resDf, sig = sig, up = up, dn = dn)
}

save_deg_results <- function(res, prefix, outDir) {
  write.table(res$all, file = file.path(outDir, paste0(prefix, "_all.txt")),
              sep = "\t", row.names = FALSE, quote = FALSE)
  write.table(res$up, file = file.path(outDir, paste0(prefix, "_up.txt")),
              sep = "\t", row.names = FALSE, quote = FALSE)
  write.table(res$dn, file = file.path(outDir, paste0(prefix, "_down.txt")),
              sep = "\t", row.names = FALSE, quote = FALSE)
}

prepare_logfc_distribution_df <- function(df, label, padjCut = 0.05, logFCCut = 0.26) {
  df$significance <- "Not significant"
  df$significance[df$adj.P.Val <= padjCut & df$logFC >= logFCCut] <- "Up"
  df$significance[df$adj.P.Val <= padjCut & df$logFC <= -logFCCut] <- "Down"
  df$significance <- factor(df$significance, levels = c("Down", "Not significant", "Up"))
  df$comparison <- label
  df
}

prepare_pvalue_distribution_df <- function(df, label, padjCut = 0.05, logFCCut = 0.26) {
  df$significance <- "Not significant"
  df$significance[df$adj.P.Val <= padjCut & df$logFC >= logFCCut] <- "Up"
  df$significance[df$adj.P.Val <= padjCut & df$logFC <= -logFCCut] <- "Down"
  df$significance <- factor(df$significance, levels = c("Down", "Not significant", "Up"))
  df$comparison <- label
  df
}

run_go_bp <- function(gene_df, title, out_prefix, outDir) {
  if (is.null(gene_df) || nrow(gene_df) == 0) {
    cat("Skipping", title, "- no significant genes.\n")
    return(invisible(NULL))
  }

  gene_ids <- unique(na.omit(gene_df$ensid))
  if (length(gene_ids) == 0) {
    cat("Skipping", title, "- no Ensembl IDs available for enrichment.\n")
    return(invisible(NULL))
  }

  go_obj <- enrichGO(
    gene = gene_ids,
    OrgDb = org.Hs.eg.db,
    ont = "BP",
    keyType = "ENSEMBL",
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05
  )

  go_df <- data.frame(go_obj)
  if (is.null(go_df) || nrow(go_df) == 0) {
    cat("Skipping", title, "- enrichGO returned no terms.\n")
    return(invisible(NULL))
  }

  p_go <- dotplot(go_obj, showCategory = 15, font.size = 8) +
    ggtitle(title)
  ggsave(file.path(outDir, paste0(out_prefix, ".pdf")), p_go, width = 8, height = 6)
  write.table(go_df, file = file.path(outDir, paste0(out_prefix, ".txt")),
              sep = "\t", row.names = FALSE, quote = FALSE)

  invisible(go_obj)
}
```

## Model 1: ~ diag + centered WM score

Uses all samples. The WM score (`RF_score_clus2`) captures the GM/WM tissue composition as a continuous covariate. We center the WM score so the `diagAD` coefficient is interpretable as the AD vs N effect at the average WM score.

```{r model1-wm-score}
tmpDf <- metaDf
tmpDf <- tmpDf[order(tmpDf$diag, decreasing = TRUE), ]
tmpDf$diag <- factor(tmpDf$diag, levels = c("N", "AD"))
tmpDf$WM_score_c <- as.numeric(scale(tmpDf$RF_score_clus2, center = TRUE, scale = FALSE))

design1 <- model.matrix(~ diag + WM_score_c, data = tmpDf)
rownames(design1) <- tmpDf$newId
head(design1)

countDf1 <- countDf[, match(rownames(design1), colnames(countDf))]
stopifnot(all(colnames(countDf1) == rownames(design1)))

cat("Model 1: ~ diag + WM_score_c\n")
res_m1 <- run_limma(countDf1, design1, coef = "diagAD", geneIDNameDf)
save_deg_results(res_m1, "DEG_model1_diag_WMscore", outDir)
```

## Compare with original stratified DEGs

Load the original stratified DEG results (from `find_degs_limma.R`) and compare overlap.

```{r compare-stratified}
origDir <- "/home/zhijiany/workdir/2026/0402_AD_PaperReviews/data/AD/DEG_per_subtype"

# Read original stratified results if available
deg_13_file <- file.path(origDir, "DEG_1_3_ADvsN_ROSMAP.txt")
deg_2_file <- file.path(origDir, "DEG_2_ADvsN_ROSMAP.txt")

if (file.exists(deg_13_file) && file.exists(deg_2_file)) {
  deg_13_orig <- read.table(deg_13_file, header = TRUE, sep = "\t",
                            stringsAsFactors = FALSE)
  deg_2_orig <- read.table(deg_2_file, header = TRUE, sep = "\t",
                           stringsAsFactors = FALSE)

  # Filter to significant DEGs with same thresholds
  padjCut <- 0.05
  logFCCut <- 0.26

  sig_13 <- deg_13_orig[deg_13_orig$adj.P.Val <= padjCut &
                          abs(deg_13_orig$logFC) >= logFCCut, "ensid"]
  sig_2 <- deg_2_orig[deg_2_orig$adj.P.Val <= padjCut &
                        abs(deg_2_orig$logFC) >= logFCCut, "ensid"]
  sig_orig_union <- union(sig_13, sig_2)

  sig_m1 <- union(res_m1$up$ensid, res_m1$dn$ensid)

  cat("Original stratified approach (adj.P <= 0.05 & |logFC| >= 0.26):\n")
  cat("  GM (clus 1&3) DEGs:", length(sig_13), "\n")
  cat("  WM (clus 2) DEGs:", length(sig_2), "\n")
  cat("  Union:", length(sig_orig_union), "\n\n")
  cat("Covariate model (adj.P <= 0.05 & |logFC| >= 0.26):\n")
  cat("  Model 1 (~ diag + WM_score_c), diagAD:", length(sig_m1), "\n")

  # Venn: Model 1 vs original union
  pdf(file.path(outDir, "venn_stratified_vs_covariate.pdf"), width = 6, height = 5)
  grid.draw(venn.diagram(
    x = list("Stratified (union)" = sig_orig_union,
             "Covariate (diag+WM)" = sig_m1),
    filename = NULL,
    fill = c("#B0AECC", "#B5A054"), alpha = 0.5,
    main = "Stratified vs Covariate Model",
    cat.pos = c(-20, 20), cat.dist = c(0.05, 0.05)
  ))
  dev.off()

  # Three-way: stratified GM, stratified WM, covariate model
  pdf(file.path(outDir, "venn_subtypes_vs_covariate.pdf"), width = 6, height = 5)
  grid.draw(venn.diagram(
    x = list("Strat GM (1&3)" = sig_13,
             "Strat WM (2)" = sig_2,
             "Covariate" = sig_m1),
    filename = NULL,
    fill = c("#B0AECC", "#7EC8B0", "#B5A054"), alpha = 0.5,
    main = "Stratified subtypes vs Covariate Model"
  ))
  dev.off()

} else {
  cat("Original stratified DEG files not found. Skipping comparison.\n")
}
```

## Volcano plots

```{r volcano-plots, fig.width=10, fig.height=6}
library(ggrepel)

make_volcano <- function(resDf, title, padjCut = 0.05, logFCCut = 0.26, nLabel = 15) {
  resDf$sig <- "NS"
  resDf$sig[resDf$adj.P.Val <= padjCut & resDf$logFC >= logFCCut] <- "Up"
  resDf$sig[resDf$adj.P.Val <= padjCut & resDf$logFC <= -logFCCut] <- "Down"
  resDf$sig <- factor(resDf$sig, levels = c("Down", "NS", "Up"))

  # Select top genes to label: highest -log10(adj.P.Val) among significant
  sigGenes <- resDf[resDf$sig %in% c("Up", "Down"), ]
  topGenes <- head(sigGenes[order(sigGenes$adj.P.Val), ], nLabel)
  resDf$display_name <- ifelse(
    is.na(resDf$symbol) | trimws(resDf$symbol) == "" | tolower(trimws(resDf$symbol)) == "na",
    resDf$ensid,
    resDf$symbol
  )
  resDf$label <- ifelse(resDf$ensid %in% topGenes$ensid, resDf$display_name, NA)

  ggplot(resDf, aes(x = logFC, y = -log10(adj.P.Val), color = sig)) +
    geom_point(size = 0.5, alpha = 0.5) +
    scale_color_manual(values = c("Down" = "blue", "NS" = "grey70", "Up" = "red")) +
    geom_hline(yintercept = -log10(padjCut), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = c(-logFCCut, logFCCut), linetype = "dashed", color = "grey40") +
    geom_text_repel(aes(label = label), size = 3, max.overlaps = 20,
                    show.legend = FALSE) +
    labs(title = title, x = "log2 Fold Change", y = "-log10(adj. P-value)") +
    theme_bw() +
    theme(legend.position = "bottom")
}

p <- make_volcano(res_m1$all, "~ diag + WM score")
ggsave(file.path(outDir, "volcano_diag_WMscore.pdf"), p, width = 7, height = 5)
p
```

## logFC distribution from saved covariate-model results

```{r logfc-distribution, fig.width=10, fig.height=4.5}
padjCut <- 0.05
logFCCut <- 0.26

degFile <- file.path(outDir, "DEG_model1_diag_WMscore_all.txt")
deg_logfc_df <- read.table(
  degFile,
  header = TRUE,
  sep = "\t",
  stringsAsFactors = FALSE,
  check.names = FALSE
)
deg_logfc_df <- prepare_logfc_distribution_df(
  deg_logfc_df,
  "Covariate model",
  padjCut = padjCut,
  logFCCut = logFCCut
)

p_logfc_hist <- ggplot(deg_logfc_df, aes(x = logFC, fill = significance)) +
  geom_histogram(binwidth = 0.05, color = "white", alpha = 0.9) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey30") +
  scale_fill_manual(values = c("Down" = "#3B6FB6", "Not significant" = "grey80", "Up" = "#C84C3A")) +
  labs(
    title = "Distribution of logFC values from saved covariate-model results",
    subtitle = "File: DEG_model1_diag_WMscore_all.txt; colors mark DEG significance",
    x = "log2 Fold Change",
    y = "Number of genes",
    fill = NULL
  ) +
  theme_bw() +
  theme(legend.position = "bottom")

cat("Genes loaded from saved results:", nrow(deg_logfc_df), "\n")
cat("Significant DEGs in saved results:", sum(deg_logfc_df$significance %in% c("Down", "Up")), "\n")

ggsave(file.path(outDir, "logFC_distribution_model1_from_file.pdf"), p_logfc_hist,
       width = 8, height = 4.5)

p_logfc_hist
```

## Raw p-value distribution from saved covariate-model results

```{r pvalue-distribution, fig.width=10, fig.height=4.5}
pvalueFile <- file.path(outDir, "DEG_model1_diag_WMscore_all.txt")
pvalue_df <- read.table(
  pvalueFile,
  header = TRUE,
  sep = "\t",
  stringsAsFactors = FALSE,
  check.names = FALSE
)
pvalue_df <- prepare_pvalue_distribution_df(
  pvalue_df,
  "Covariate model",
  padjCut = padjCut,
  logFCCut = logFCCut
)

p_pvalue_hist <- ggplot(pvalue_df, aes(x = P.Value, fill = significance)) +
  geom_histogram(binwidth = 0.02, color = "white", alpha = 0.9, boundary = 0) +
  scale_fill_manual(values = c("Down" = "#3B6FB6", "Not significant" = "grey80", "Up" = "#C84C3A")) +
  labs(
    title = "Distribution of raw p-values from saved covariate-model results",
    subtitle = "File: DEG_model1_diag_WMscore_all.txt; colors mark DEG significance",
    x = "Raw p-value",
    y = "Number of genes",
    fill = NULL
  ) +
  theme_bw() +
  theme(legend.position = "bottom")

cat("Genes loaded from saved p-value results:", nrow(pvalue_df), "\n")
cat("Genes with raw P.Value < 0.05:", sum(pvalue_df$P.Value < 0.05), "\n")

ggsave(file.path(outDir, "pvalue_distribution_model1_from_file.pdf"), p_pvalue_hist,
       width = 8, height = 4.5)

p_pvalue_hist
```

## logFC distribution for stratified DEG result files

```{r logfc-distribution-stratified, fig.width=12, fig.height=4.8}
deg_13_file <- "/home/zhijiany/workdir/2026/0402_AD_PaperReviews/data/AD/DEG_per_subtype/DEG_1_3_ADvsN_ROSMAP.txt"
deg_2_file <- "/home/zhijiany/workdir/2026/0402_AD_PaperReviews/data/AD/DEG_per_subtype/DEG_2_ADvsN_ROSMAP.txt"

deg_13_df <- read.table(
  deg_13_file,
  header = TRUE,
  sep = "\t",
  stringsAsFactors = FALSE,
  check.names = FALSE
)
deg_2_df <- read.table(
  deg_2_file,
  header = TRUE,
  sep = "\t",
  stringsAsFactors = FALSE,
  check.names = FALSE
)

deg_13_df <- prepare_logfc_distribution_df(
  deg_13_df,
  "Stratified GM (clus 1&3)",
  padjCut = padjCut,
  logFCCut = logFCCut
)
deg_2_df <- prepare_logfc_distribution_df(
  deg_2_df,
  "Stratified WM (clus 2)",
  padjCut = padjCut,
  logFCCut = logFCCut
)

deg_stratified_logfc_df <- rbind(deg_13_df, deg_2_df)

cat("Genes loaded from DEG_1_3_ADvsN_ROSMAP.txt:", nrow(deg_13_df), "\n")
cat("Significant DEGs in DEG_1_3_ADvsN_ROSMAP.txt:",
    sum(deg_13_df$significance %in% c("Down", "Up")), "\n")
cat("Genes loaded from DEG_2_ADvsN_ROSMAP.txt:", nrow(deg_2_df), "\n")
cat("Significant DEGs in DEG_2_ADvsN_ROSMAP.txt:",
    sum(deg_2_df$significance %in% c("Down", "Up")), "\n")

p_logfc_stratified <- ggplot(deg_stratified_logfc_df, aes(x = logFC, fill = significance)) +
  geom_histogram(binwidth = 0.05, color = "white", alpha = 0.9) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey30") +
  facet_wrap(~comparison, ncol = 2) +
  scale_fill_manual(values = c("Down" = "#3B6FB6", "Not significant" = "grey80", "Up" = "#C84C3A")) +
  labs(
    title = "Distribution of logFC values for stratified DEG result files",
    subtitle = "Files: DEG_1_3_ADvsN_ROSMAP.txt and DEG_2_ADvsN_ROSMAP.txt",
    x = "log2 Fold Change",
    y = "Number of genes",
    fill = NULL
  ) +
  theme_bw() +
  theme(legend.position = "bottom")

ggsave(file.path(outDir, "logFC_distribution_stratified_models.pdf"), p_logfc_stratified,
       width = 10, height = 4.8)

p_logfc_stratified
```

## Raw p-value distribution for stratified DEG result files

```{r pvalue-distribution-stratified, fig.width=12, fig.height=4.8}
deg_13_pval_df <- prepare_pvalue_distribution_df(
  deg_13_df,
  "Stratified GM (clus 1&3)",
  padjCut = padjCut,
  logFCCut = logFCCut
)
deg_2_pval_df <- prepare_pvalue_distribution_df(
  deg_2_df,
  "Stratified WM (clus 2)",
  padjCut = padjCut,
  logFCCut = logFCCut
)

deg_stratified_pval_df <- rbind(deg_13_pval_df, deg_2_pval_df)

p_pvalue_stratified <- ggplot(deg_stratified_pval_df, aes(x = P.Value, fill = significance)) +
  geom_histogram(binwidth = 0.02, color = "white", alpha = 0.9, boundary = 0) +
  facet_wrap(~comparison, ncol = 2) +
  scale_fill_manual(values = c("Down" = "#3B6FB6", "Not significant" = "grey80", "Up" = "#C84C3A")) +
  labs(
    title = "Distribution of raw p-values for stratified DEG result files",
    subtitle = "Files: DEG_1_3_ADvsN_ROSMAP.txt and DEG_2_ADvsN_ROSMAP.txt",
    x = "Raw p-value",
    y = "Number of genes",
    fill = NULL
  ) +
  theme_bw() +
  theme(legend.position = "bottom")

cat("Genes with raw P.Value < 0.05 in DEG_1_3_ADvsN_ROSMAP.txt:",
    sum(deg_13_pval_df$P.Value < 0.05), "\n")
cat("Genes with raw P.Value < 0.05 in DEG_2_ADvsN_ROSMAP.txt:",
    sum(deg_2_pval_df$P.Value < 0.05), "\n")

ggsave(file.path(outDir, "pvalue_distribution_stratified_models.pdf"), p_pvalue_stratified,
       width = 10, height = 4.8)

p_pvalue_stratified
```

## Effect of WM score on gene expression (Model 1)

Extract the coefficient for `WM_score_c` to see which genes are most associated with WM composition in the additive model.

```{r wm-score-effect}
fit_wm <- lmFit(countDf1, design1)
fit_wm <- eBayes(fit_wm)
fitRes_wm <- topTable(fit_wm, coef = "WM_score_c", adjust = "BH", n = Inf)

resDf_wm <- data.frame(ensid = rownames(fitRes_wm), fitRes_wm, check.names = FALSE)
resDf_wm <- merge(resDf_wm, geneIDNameDf, by.x = "ensid", by.y = "ensid", all.x = TRUE)
resDf_wm <- resDf_wm[order(resDf_wm$adj.P.Val), ]

cat("Genes significantly associated with WM score (adj.P <= 0.05):",
    sum(resDf_wm$adj.P.Val <= 0.05), "\n")

write.table(resDf_wm, file = file.path(outDir, "WM_score_effect_all.txt"),
            sep = "\t", row.names = FALSE, quote = FALSE)

head(resDf_wm, 20)
```

## GO enrichment analysis on covariate-adjusted DEGs

```{r go-enrichment, fig.width=10, fig.height=8}
library(clusterProfiler)
library(org.Hs.eg.db)

# Use Ensembl IDs directly so DEGs with missing symbols are retained.
sig_genes <- rbind(res_m1$up, res_m1$dn)

go_all <- run_go_bp(sig_genes, "GO BP: All covariate-adjusted DEGs",
                    "GO_BP_all_DEGs", outDir)
go_up <- run_go_bp(res_m1$up, "GO BP: Up-regulated DEGs (AD > N)",
                   "GO_BP_up_DEGs", outDir)
go_dn <- run_go_bp(res_m1$dn, "GO BP: Down-regulated DEGs (AD < N)",
                   "GO_BP_down_DEGs", outDir)
```

## Summary table

```{r summary}
summary_df <- data.frame(
  Analysis = c(
    "AD vs N DEG contrast (~ diag + WM_score_c; diagAD coef)",
    "WM-score association (~ diag + WM_score_c; WM_score_c coef)"
  ),
  Significant = c(
    nrow(res_m1$sig),
    sum(resDf_wm$adj.P.Val <= 0.05)
  ),
  Positive = c(
    nrow(res_m1$up),
    sum(resDf_wm$adj.P.Val <= 0.05 & resDf_wm$logFC >= 0.26)
  ),
  Negative = c(
    nrow(res_m1$dn),
    sum(resDf_wm$adj.P.Val <= 0.05 & resDf_wm$logFC <= -0.26)
  )
)

cat("Note: the WM-score row summarizes genes associated with WM_score_c, not an AD-vs-N DEG contrast.\n")
print(summary_df)

write.table(summary_df, file = file.path(outDir, "DEG_summary.txt"),
            sep = "\t", row.names = FALSE, quote = FALSE)
```
