

Calculate the correlation between RF classification probability for cluster 2 and WM scores using enriched WM siganture genes from supplementary table S4B, with those having t stats > 10

```{r setup}
rm(list=ls())
library(readxl)
library(GSVA)
library(ggplot2)
library(dplyr)

outDir <- "/home/zhijiany/workdir/2026/0402_AD_PaperReviews/out/4.ST_WM_GSVA"
dir.create(outDir, recursive = TRUE, showWarnings = FALSE)

suppFile <- "/home/zhijiany/workdir/2026/0402_AD_PaperReviews/data/supplementary_table.xlsx"
countFile <- "/home/zhijiany/workdir/2026/0402_AD_PaperReviews/data/AD/ROSMAP_AD-N/ROSMAP_AMPAD_AD-N_counts.txt"
rfScoreFile <- "/home/zhijiany/workdir/2026/0402_AD_PaperReviews/out/2.RF_score/ROSMAP_AMPAD_AD-N_meta_with_RF_scores.txt"
rfValidationFile <- "/home/zhijiany/workdir/2026/0402_AD_PaperReviews/out/2.RF_score/validation_summary.csv"
rfPerclassFile <- "/home/zhijiany/workdir/2026/0402_AD_PaperReviews/out/2.RF_score/perclass_metrics.csv"

theme_set(theme_bw())

# cluster colors: 1=green, 2=blue, 3=red
clus_colors <- c("1" = "#00BA38", "2" = "#619CFF", "3" = "#F8766D")
```

## Helper functions

```{r helpers}
run_ssgsea_safe <- function(expr_mat, gene_sets) {
  if ("ssgseaParam" %in% getNamespaceExports("GSVA")) {
    param <- GSVA::ssgseaParam(exprData = expr_mat, geneSets = gene_sets,
                               alpha = 0.25,
                               normalize = TRUE)
    ssgsea_res <- GSVA::gsva(param, verbose = FALSE)
  } else {
    ssgsea_res <- GSVA::gsva(expr = expr_mat, gset.idx.list = gene_sets,
                             method = "ssgsea",
                             ssgsea.norm = TRUE,
                             verbose = FALSE)
  }

  as.numeric(ssgsea_res[1, ])
}

compute_correlations <- function(df, cohort_name) {
  tmp <- df |>
    filter(!is.na(RF_score_clus2), !is.na(WM_ssGSEA_score))

  if (nrow(tmp) < 3 ||
      length(unique(tmp$RF_score_clus2)) < 2 ||
      length(unique(tmp$WM_ssGSEA_score)) < 2) {
    return(data.frame(
      cohort = cohort_name,
      n = nrow(tmp),
      spearman_rho = NA_real_,
      spearman_p = NA_real_,
      pearson_r = NA_real_,
      pearson_p = NA_real_,
      stringsAsFactors = FALSE
    ))
  }

  spearman_test <- cor.test(tmp$RF_score_clus2, tmp$WM_ssGSEA_score,
                            method = "spearman", exact = FALSE)
  pearson_test <- cor.test(tmp$RF_score_clus2, tmp$WM_ssGSEA_score,
                           method = "pearson")

  data.frame(
    cohort = cohort_name,
    n = nrow(tmp),
    spearman_rho = unname(spearman_test$estimate),
    spearman_p = spearman_test$p.value,
    pearson_r = unname(pearson_test$estimate),
    pearson_p = pearson_test$p.value,
    stringsAsFactors = FALSE
  )
}

fmt_p <- function(x) {
  ifelse(x < 2.2e-16, "<2.2e-16", formatC(x, format = "e", digits = 2))
}

compute_boxplot_stats <- function(df) {
  plot_df <- df |>
    filter(!is.na(RF_predicted_clus), !is.na(WM_ssGSEA_score)) |>
    mutate(RF_predicted_clus = factor(RF_predicted_clus))

  kw_test <- kruskal.test(WM_ssGSEA_score ~ RF_predicted_clus, data = plot_df)

  group_levels <- levels(plot_df$RF_predicted_clus)
  pair_df <- utils::combn(group_levels, 2, simplify = FALSE) |>
    lapply(function(pair) {
      pair_data <- plot_df |>
        filter(RF_predicted_clus %in% pair)
      pair_test <- wilcox.test(WM_ssGSEA_score ~ RF_predicted_clus, data = pair_data,
                               exact = FALSE)
      data.frame(
        group1 = pair[[1]],
        group2 = pair[[2]],
        p_value = pair_test$p.value,
        stringsAsFactors = FALSE
      )
    }) |>
    bind_rows() |>
    mutate(
      xmin = match(group1, group_levels),
      xmax = match(group2, group_levels)
    )

  y_max <- max(plot_df$WM_ssGSEA_score, na.rm = TRUE)
  y_min <- min(plot_df$WM_ssGSEA_score, na.rm = TRUE)
  y_range <- y_max - y_min
  if (!is.finite(y_range) || y_range == 0) {
    y_range <- max(abs(y_max), 1)
  }
  step_height <- 0.08 * y_range
  tick_height <- 0.015 * y_range
  label_offset <- 0.02 * y_range

  pair_df <- pair_df |>
    arrange(xmin, xmax) |>
    mutate(
      y = y_max + step_height * seq_len(n()),
      tick_y = y - tick_height,
      label_y = y + label_offset,
      label = paste0("Wilcoxon p=", fmt_p(p_value))
    )

  list(
    plot_df = plot_df,
    kw_label = paste0(
      "Kruskal-Wallis chi-squared=", sprintf("%.2f", unname(kw_test$statistic)),
      ", p=", fmt_p(kw_test$p.value)
    ),
    y_min = y_min,
    kw_y = y_max + step_height * (nrow(pair_df) + 1.2),
    y_upper = y_max + step_height * (nrow(pair_df) + 2.2),
    pair_df = pair_df
  )
}
```

## Load RF-scored metadata and counts

```{r load-data}
metaDf <- read.table(rfScoreFile, header = TRUE, sep = "\t",
                     stringsAsFactors = FALSE, check.names = FALSE)
metaDf$sample_id <- sub("^X", "", metaDf$newId)

countDf <- read.table(countFile, header = TRUE, sep = "\t",
                      stringsAsFactors = FALSE, check.names = FALSE)
rownames(countDf) <- countDf[, 1]
countDf <- as.matrix(countDf[, -1])
storage.mode(countDf) <- "numeric"

stopifnot(all(metaDf$sample_id %in% colnames(countDf)))
metaDf <- metaDf[match(colnames(countDf), metaDf$sample_id), ]
stopifnot(all(metaDf$sample_id == colnames(countDf)))

cat("Samples in metadata:", nrow(metaDf), "\n")
cat("Samples in counts:", ncol(countDf), "\n")
print(table(metaDf$study))
```

## Load WM signature from Supplementary Table S4B

```{r signature}
wmSigRaw <- read_excel(suppFile, sheet = "Table S4B")
wmSigRaw <- as.data.frame(wmSigRaw, stringsAsFactors = FALSE)

wmSigDf <- wmSigRaw |>
  select(ensembl, gene, t_stat_WM, p_value_WM, fdr_WM) |>
  filter(!is.na(ensembl), !is.na(t_stat_WM), t_stat_WM > 10) |>
  arrange(desc(t_stat_WM))

wmGenesAll <- unique(wmSigDf$ensembl)
wmGenesPresent <- intersect(wmGenesAll, rownames(countDf))
wmGenesMissing <- setdiff(wmGenesAll, rownames(countDf))

cat("WM signature genes (t_stat_WM > 10):", length(wmGenesAll), "\n")
cat("Present in counts:", length(wmGenesPresent), "\n")
cat("Missing from counts:", length(wmGenesMissing), "\n")

stopifnot(length(wmGenesAll) == 302)
stopifnot(length(wmGenesPresent) == 299)
stopifnot(length(wmGenesMissing) == 3)

wmSigPresentDf <- wmSigDf |> filter(ensembl %in% wmGenesPresent)
wmSigMissingDf <- wmSigDf |> filter(ensembl %in% wmGenesMissing)

write.table(wmSigPresentDf,
            file = file.path(outDir, "WM_signature_genes_tstat_gt10_present.txt"),
            sep = "\t", row.names = FALSE, quote = FALSE)
write.table(wmSigMissingDf,
            file = file.path(outDir, "WM_signature_genes_tstat_gt10_missing.txt"),
            sep = "\t", row.names = FALSE, quote = FALSE)

summaryLines <- c(
  paste("WM signature genes with t_stat_WM > 10:", length(wmGenesAll)),
  paste("Present in counts:", length(wmGenesPresent)),
  paste("Missing from counts:", length(wmGenesMissing)),
  paste("Missing genes:", paste(wmSigMissingDf$gene, collapse = ", "))
)
writeLines(summaryLines, con = file.path(outDir, "WM_signature_summary.txt"))
wmSigMissingDf
```

## Prepare expression matrix for ssGSEA

The ROSMAP matrix is already continuous normalized expression, so we use it
directly for ssGSEA rather than re-running count-based normalization. Because
ssGSEA is based on within-sample gene ranks, we score the combined matrix once
across all samples and keep the default ssGSEA normalization enabled
(`normalize = TRUE` or `ssgsea.norm = TRUE`) so WM enrichment scores are on a
more comparable cross-sample scale for visualization and downstream summaries.

```{r expression-matrix}
exprMat <- countDf

stopifnot(identical(colnames(exprMat), metaDf$sample_id))
stopifnot(identical(rownames(exprMat), rownames(countDf)))

normSummaryDf <- metaDf |>
  count(study, name = "n_samples")
write.table(normSummaryDf, file = file.path(outDir, "cohort_sample_counts.txt"),
            sep = "\t", row.names = FALSE, quote = FALSE)

summary(as.numeric(exprMat[, 1:5]))
```

## Compute WM ssGSEA score per sample

```{r ssgsea-score}
wmScores <- run_ssgsea_safe(
  expr_mat = exprMat,
  gene_sets = list(WM_signature = wmGenesPresent)
)

gsvaDf <- data.frame(
  sample_id = colnames(exprMat),
  WM_ssGSEA_score = wmScores,
  stringsAsFactors = FALSE
)

resDf <- metaDf |>
  left_join(gsvaDf, by = "sample_id")

stopifnot(sum(is.na(resDf$WM_ssGSEA_score)) == 0)
stopifnot(sum(is.na(resDf$RF_score_clus2)) == 0)

write.table(resDf,
            file = file.path(outDir, "ROSMAP_AMPAD_AD-N_meta_with_RF_and_ssGSEA_scores.txt"),
            sep = "\t", row.names = FALSE, quote = FALSE)

summary(resDf$WM_ssGSEA_score)
```

## Correlation between ssGSEA WM score and RF cluster 2 score

```{r correlations}
corSummaryDf <- bind_rows(
  compute_correlations(resDf, "Pooled"),
  compute_correlations(filter(resDf, study == "MAP"), "MAP"),
  compute_correlations(filter(resDf, study == "ROS"), "ROS")
)

corSummaryDf$label <- paste0(
  corSummaryDf$cohort,
  ": rho=", sprintf("%.3f", corSummaryDf$spearman_rho),
  ", p=", fmt_p(corSummaryDf$spearman_p),
  ", n=", corSummaryDf$n
)

write.table(corSummaryDf,
            file = file.path(outDir, "WM_ssGSEA_vs_RF_score_correlations.txt"),
            sep = "\t", row.names = FALSE, quote = FALSE)

corSummaryDf
```

## Reviewer-facing figures

```{r figures}
pooled_label <- corSummaryDf$label[corSummaryDf$cohort == "Pooled"]

p_scatter <- ggplot(resDf, aes(x = RF_score_clus2, y = WM_ssGSEA_score, color = study)) +
  geom_point(size = 2.2, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  geom_smooth(
    data = resDf,
    aes(x = RF_score_clus2, y = WM_ssGSEA_score),
    inherit.aes = FALSE,
    method = "lm", se = FALSE, color = "black",
    linetype = "dashed", linewidth = 0.9
  ) +
  annotate("text", x = Inf, y = -Inf, label = pooled_label,
           hjust = 1.05, vjust = -0.8, size = 3.7) +
  labs(
    title = "WM ssGSEA score tracks RF cluster 2 score",
    subtitle = "Colored lines: cohort-specific fits; dashed black line: pooled fit",
    x = "RF_score_clus2",
    y = "WM ssGSEA score",
    color = "Cohort"
  )

ggsave(file.path(outDir, "WM_ssGSEA_vs_RF_score_scatter.pdf"),
       p_scatter, width = 5.8, height = 5.8)

p_dist <- ggplot(resDf, aes(x = study, y = WM_ssGSEA_score, fill = study)) +
  geom_violin(trim = FALSE, alpha = 0.5, color = NA) +
  geom_boxplot(width = 0.18, outlier.shape = NA, alpha = 0.85) +
  labs(
    title = "Distribution of WM ssGSEA scores by cohort",
    x = "Cohort",
    y = "WM ssGSEA score",
    fill = "Cohort"
  ) +
  guides(fill = "none")

ggsave(file.path(outDir, "WM_ssGSEA_score_by_cohort.pdf"),
       p_dist, width = 5.5, height = 5)

clusterStatList <- compute_boxplot_stats(resDf)

p_cluster <- ggplot(clusterStatList$plot_df,
                    aes(x = RF_predicted_clus, y = WM_ssGSEA_score,
                        fill = RF_predicted_clus)) +
  geom_boxplot(alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.12, alpha = 0.35, size = 1.2) +
  geom_segment(data = clusterStatList$pair_df,
               aes(x = xmin, xend = xmax, y = y, yend = y),
               inherit.aes = FALSE, linewidth = 0.5) +
  geom_segment(data = clusterStatList$pair_df,
               aes(x = xmin, xend = xmin, y = tick_y, yend = y),
               inherit.aes = FALSE, linewidth = 0.5) +
  geom_segment(data = clusterStatList$pair_df,
               aes(x = xmax, xend = xmax, y = tick_y, yend = y),
               inherit.aes = FALSE, linewidth = 0.5) +
  geom_text(data = clusterStatList$pair_df,
            aes(x = (xmin + xmax) / 2, y = label_y, label = label),
            inherit.aes = FALSE, size = 3.1) +
  annotate("text", x = 2, y = clusterStatList$kw_y, label = clusterStatList$kw_label,
           size = 3.3) +
  coord_cartesian(ylim = c(clusterStatList$y_min, clusterStatList$y_upper)) +
  scale_fill_manual(values = clus_colors) +
  labs(
    title = "WM ssGSEA score by RF-predicted cluster",
    x = "RF-predicted cluster",
    y = "WM ssGSEA score",
    fill = "RF cluster"
  )

ggsave(file.path(outDir, "WM_ssGSEA_score_by_RF_predicted_cluster.pdf"),
       p_cluster, width = 6.2, height = 5)

p_scatter
```

## Existing RF validation metrics

```{r rf-validation}
rfValidationDf <- read.csv(rfValidationFile, stringsAsFactors = FALSE)
rfPerclassDf <- read.csv(rfPerclassFile, stringsAsFactors = FALSE)

write.table(rfValidationDf,
            file = file.path(outDir, "RF_validation_summary_reused.txt"),
            sep = "\t", row.names = FALSE, quote = FALSE)
write.table(rfPerclassDf,
            file = file.path(outDir, "RF_perclass_metrics_reused.txt"),
            sep = "\t", row.names = FALSE, quote = FALSE)

rfValidationDf
```

```{r}
rfPerclassDf
```
