

Run DASC with different parameters to confirm the robustness of the algorithm. 
Original run information: Consensus clustering matrix, cophenetic coefficient value, and silhouette value computed over 500 iterations at k = 2-8 by the DASC algorithm.

- random initiation
- different numbers of clusters
- ARI distribution plot for different parameters

Need:
1. DASC package
    - https://academic.oup.com/bioinformatics/article/34/7/1141/4386916
    - https://github.com/zhanglabNKU/DASC
    - installed in DASC conda env
    
2. ROSMAP data
    - 
3. CZ's script
    - /home/zhijiany/workdir/2026/0402_AD_PaperReviews/data/AD/run_DASC.R
4. Understand ARI


```{r}
rm(list=ls())

#load_lib
#load the libraries
#rm(list=ls())
library(dplyr)
library(ggplot2)
#library('DESeq2')
#library(Rtsne)
library(CancerSubtypes)
library(NMF)
#library(sva)
#library(GGally)
library(DASC)
#library(SNFtool)
#library(ggpubr)
#library('limma')
library(pheatmap)
library(RColorBrewer)
#library(factoextra)
#require(cowplot)
library(parallel)
library(mclust)

# Output directory
outDir = '/home/zhijiany/workdir/2026/0402_AD_PaperReviews/out/1.DASC_multipleRun'
dir.create(outDir, recursive = TRUE, showWarnings = FALSE)
```

# MAP DASC test run
```{r}

# load expression data
countTableSel <- read.table(
"/home/zhijiany/workdir/2026/0402_AD_PaperReviews/data/AD/ROSMAP_AD-N/ROSMAP_AMPAD_AD-N_counts.txt",
  header = TRUE, sep = '\t', stringsAsFactors = FALSE
)
rownames(countTableSel) <- countTableSel[, 1]
countTableSel <- countTableSel[, -1]
colnames(countTableSel) <- gsub('^X', '', colnames(countTableSel))

# load meta data
metaTableSel <- read.table(
  "/home/zhijiany/workdir/2026/0402_AD_PaperReviews/data/AD/ROSMAP_AD-N/ROSMAP_AMPAD_AD-N_meta.txt",
  header = TRUE, sep = '\t', stringsAsFactors = FALSE
)

# filter to MAP study, AD only (diag == 'AD', cogdx == 4)
metaTableSel <- metaTableSel[metaTableSel$study == 'MAP' & metaTableSel$diag == 'AD', ]
countTableSel <- countTableSel[, colnames(countTableSel) %in% metaTableSel$newId]
metaTableSel <- metaTableSel[match(colnames(countTableSel), metaTableSel$newId), ]
cat('MAP AD samples:', ncol(countTableSel), '\n')  # should be 68

# verify samples match
all(colnames(countTableSel) %in% metaTableSel$newId)
all(colnames(countTableSel) == metaTableSel$newId)

# obtain pdata
selVec <- c('study', 'educ', 'age_death', 'msex', 'braaksc', 'ceradsc', 'cogdx', 'Batch', 'newId')
pdata <- dplyr::select(metaTableSel, selVec)
rownames(pdata) <- pdata$newId

# verify data alignment
all(pdata$newId == colnames(countTableSel))

# select top3000 genes by MAD
topVal <- 3000                                            
subMatrix_top_MAD <- FSbyMAD(countTableSel, cut.type = "topk", value = topVal)

# shift if negative values exist since NMF requires input >=0
if (min(subMatrix_top_MAD) < 0) {
  subMatrix_top_MAD <- subMatrix_top_MAD + abs(min(subMatrix_top_MAD))
}

# setup DASC parameters
minRank <- 2                                                                                        
maxRank <- 8
nRun <- 10                                             
               
# run DASC
start.time <- Sys.time()
res <- DASC(edata = subMatrix_top_MAD, pdata = pdata,
            factor = pdata$msex,
            method = 'ama', type = 3, lambda = 0.001, rank = minRank:maxRank,
            nrun = nRun, annotation = 'rosmap')
end.time <- Sys.time()
time.taken <- end.time - start.time
cat('time:\n')
print(time.taken)


```

# DASC cluster heatmap test

```{r}       
# Pick rank based on your rank survey plot                                                  
rankSel <- 3                                                                                
                                                                                          
# Consensus heatmap for that rank                                                           
pdf(file.path(outDir, paste0('DASC_consensus_k', rankSel, '.pdf')))
consensusmap(res[[as.character(rankSel)]]$consensus)                      
dev.off()  

```

# Sensitivity analysis — parameter grid

```{r}
# Parameter grid: topVal x lambda x nRun
topVals   <- c(2000, 3000, 4000)
lambdas   <- c(0.001, 0.01, 0.1)
nRunVals  <- c(10, 100, 500)
minRank   <- 2
maxRank   <- 8

# Output subdirectory for sensitivity runs
sensDir <- file.path(outDir, 'sensitivity')
dir.create(sensDir, recursive = TRUE, showWarnings = FALSE)

# Pre-compute MAD matrices for each topVal (avoid redundant computation)
madMatrices <- list()
for (tv in topVals) {
  mat <- FSbyMAD(countTableSel, cut.type = "topk", value = tv)
  if (min(mat) < 0) {
    mat <- mat + abs(min(mat))
  }
  madMatrices[[as.character(tv)]] <- mat
}
```

```{r}
# Run DASC across all parameter combinations, looping over nRun values
paramGrid <- expand.grid(topVal = topVals, lambda = lambdas,
                         stringsAsFactors = FALSE)

# Store results per nRun: allResultsByNrun[["nrun10"]] <- list of 9 DASC results
allResultsByNrun <- list()

for (nr in nRunVals) {
  nrLabel <- paste0('nrun', nr)
  nrDir   <- file.path(sensDir, nrLabel)
  dir.create(nrDir, recursive = TRUE, showWarnings = FALSE)

  cat('\n##############################\n')
  cat('# nRun =', nr, '\n')
  cat('##############################\n')

  resultsList <- list()

  for (i in seq_len(nrow(paramGrid))) {
    tv  <- paramGrid$topVal[i]
    lam <- paramGrid$lambda[i]

    runLabel <- paste0('top', tv, '_lambda', lam)
    cat('\n=== Run', i, '/', nrow(paramGrid), ':', runLabel, '(nRun =', nr, ') ===\n')

    start.time <- Sys.time()
    res_i <- DASC(edata = madMatrices[[as.character(tv)]],
                  pdata = pdata,
                  factor = pdata$msex,
                  method = 'ama', type = 3, lambda = lam,
                  rank = minRank:maxRank,
                  nrun = nr, annotation = 'rosmap')
    elapsed <- difftime(Sys.time(), start.time, units = 'mins')
    cat('  Finished in', round(as.numeric(elapsed), 1), 'min\n')

    # Save individual result
    saveRDS(res_i, file = file.path(nrDir, paste0('DASC_', runLabel, '.rds')))

    resultsList[[runLabel]] <- res_i
  }

  # Save full collection for this nRun
  saveRDS(resultsList, file = file.path(nrDir, 'DASC_allResults.rds'))
  allResultsByNrun[[nrLabel]] <- resultsList
}

cat('\nAll runs complete.\n')
```

# Sensitivity analysis — ARI comparison (per nRun)

```{r}
# Extract cluster labels for each run at each rank
extractLabels <- function(allResults, rank) {
  labels <- sapply(allResults, function(res_i) {
    r <- as.character(rank)
    if (!is.null(res_i[[r]])) {
      as.integer(res_i[[r]]$class)
    } else {
      rep(NA, length(res_i[[1]]$class))
    }
  })
  labels
}

# Compute and save ARI heatmaps + summary for each nRun value
for (nr in nRunVals) {
  nrLabel <- paste0('nrun', nr)
  nrDir   <- file.path(sensDir, nrLabel)
  allResults <- allResultsByNrun[[nrLabel]]

  # --- Pairwise ARI matrices ---
  ariByRank <- list()
  for (k in minRank:maxRank) {
    labels <- extractLabels(allResults, k)
    nPairs <- ncol(labels)
    ariMat <- matrix(1, nrow = nPairs, ncol = nPairs,
                     dimnames = list(colnames(labels), colnames(labels)))
    for (i in 1:(nPairs - 1)) {
      for (j in (i + 1):nPairs) {
        ari <- adjustedRandIndex(labels[, i], labels[, j])
        ariMat[i, j] <- ari
        ariMat[j, i] <- ari
      }
    }
    ariByRank[[as.character(k)]] <- ariMat
  }
  saveRDS(ariByRank, file = file.path(nrDir, 'ARI_byRank.rds'))

  # --- ARI heatmaps per rank ---
  for (k in minRank:maxRank) {
    ariMat <- ariByRank[[as.character(k)]]
    pdf(file.path(nrDir, paste0('ARI_heatmap_rank', k, '.pdf')),
        width = 10, height = 8)
    pheatmap(ariMat,
             display_numbers = TRUE,
             number_format = "%.2f",
             cluster_rows = FALSE,
             cluster_cols = FALSE,
             main = paste0('Pairwise ARI (k = ', k, ', nRun = ', nr, ')'),
             color = colorRampPalette(c("white", "steelblue"))(100),
             fontsize_number = 7)
    dev.off()
  }

  # --- Summary: mean ARI per parameter setting at each rank ---
  ariSummary <- data.frame()
  for (k in minRank:maxRank) {
    ariMat <- ariByRank[[as.character(k)]]
    for (i in seq_len(nrow(ariMat))) {
      meanARI <- mean(ariMat[i, -i])
      ariSummary <- rbind(ariSummary, data.frame(
        rank = k,
        runLabel = rownames(ariMat)[i],
        topVal = as.numeric(gsub(".*top(\\d+).*", "\\1", rownames(ariMat)[i])),
        lambda = as.numeric(gsub(".*lambda([0-9e.-]+).*", "\\1", rownames(ariMat)[i])),
        meanARI = meanARI
      ))
    }
  }

  p1 <- ggplot(ariSummary, aes(x = factor(rank), y = meanARI, color = factor(lambda))) +
    geom_point(size = 2) +
    geom_line(aes(group = interaction(topVal, lambda)), alpha = 0.5) +
    facet_wrap(~ topVal, labeller = label_both) +
    labs(x = 'Rank (k)', y = 'Mean ARI vs other settings',
         color = 'Lambda',
         title = paste0('Subgroup stability across parameter choices (nRun = ', nr, ')')) +
    theme_bw()

  ggsave(file.path(nrDir, 'ARI_summary.pdf'), p1, width = 12, height = 6)
  cat('Saved ARI results for nRun =', nr, '\n')
}
```